---
title: "NYC Taxi and Limousine - High Volume Dashboard"
author: "John Cruz"
date: "2023-04-01"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The NYC Taxi and Limousine Commission publishes through their NYC Open Data program trip information on different designations. One such category are *High Volume For-Hire Vehicles* (HVFHV) which comprise of app based ride requests such as Uber and Lyft. The data being used was collected from their website and will be used for this analysis.

**Source: ** 

[TLC Trip Record Data](https://www.nyc.gov/site/tlc/about/tlc-trip-record-data.page)
[Taxi Zones](https://data.cityofnewyork.us/Transportation/NYC-Taxi-Zones/d3c5-ddgc)

The NYC Taxi and Limousine Commission also provides data dictionaries for their datasets. 

[HVFHV Data Dictionary](https://www.nyc.gov/assets/tlc/downloads/pdf/data_dictionary_trip_records_hvfhs.pdf)

## Objective

Create a Tableau Public dashboard that provides descriptive information based on the HVFHV industry for the month of August 2022.

## Required Libraries

To be able to read the parquet type files, you must have the *arrow* library installed. 

```{r library, message=FALSE}
library(tidyverse)
library(arrow)
```

## Import Manhattan Trips Data

```{r import-trips}
trips_df <- read_parquet('manhattan_trips.parquet')
```

## Descriptive Statistics

```{r rideshare}
requested <-
  trips_df |> 
  group_by(app, shared_request_flag) |> 
  summarise(count = n()) |> 
  mutate(rideshare = 'requested') |> 
  pivot_wider(names_from = shared_request_flag, values_from = count)

matched <-
  trips_df |> 
  group_by(app, shared_match_flag) |> 
  summarise(count = n()) |> 
  mutate(rideshare = 'matched') |> 
  pivot_wider(names_from = shared_match_flag, values_from = count)

pct_requested <-
  requested |> 
  full_join(matched) |> 
  mutate(total = N + Y) |> 
  select(rideshare, Y) |> 
  pivot_wider(names_from = rideshare, values_from = Y) |> 
  mutate(pct_requested = matched / requested)

write.table(pct_requested, file = 'rideshare.txt', row.names = FALSE)
```


